The Precambrian paleogeography of Laurentia

### 2020-11-04 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Update Laurentia_Paleogeo_Manuscript.tex

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until ca. 620 Ma north of the New York promontory.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continous with the South Australia his interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny. Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}. However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga.

Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related terranes including East Svalbard \citep{McClelland2019a}. These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}. The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation \citep{Cawood2010a}. Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g. \citealp{Evans2009a} and as shown in Fig. \ref{fig:Grenville_reconstructions}). A challenge with this model is it is unclear what was the conjugate lithosphere that rifted from East Greenland leading to thick sedimentary successions interpreted to have been deposited on a thermally-subsiding passive margin from ca. 850 Ma into the Cryogenian (Fig. \ref{fig:tectonic_history}; \citealp{Maloof2006a}). \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting. The separation of Baltica from Laurentia is hypothesized to have occurred ca. 600 Ma associated with the opening of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citep{Cawood2001a}).

@@

High-quality paleomagnetic poles constrain the coeval ca. 1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}. These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone. However, with the subsequent rapid motion of Laurentia to low latitudes (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny. This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g. \citealp{Li2008a}). Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift. The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g. \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia. The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them become conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}. The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca. 1090 followed by peak granulite metamorphism ca. 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean. If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia may have initiated ca. 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in eastern Laurentia.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} argue similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the ca. 1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. In terms of departing from this position, one possibility is that its departure is associated with early Neoproterozoic extension in northwest Laurentia.

>>>>

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until after  ca. 620 Ma north of the New York promontory.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to be the  result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continuous with the South Australia his interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny. A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup-Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia ca. 1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica \citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}. However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga. In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic mortion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig. \ref{fig:Grenville_reconstructions}).

Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related terranes including East Svalbard \citep{McClelland2019a}. These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}. The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}. Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g. \citealp{Evans2009a} and as shown in Fig. \ref{fig:Grenville_reconstructions}). This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP that would have Laurentia journey after the Grenville loop northward across the equator and back south to a similar Grenville loop position prior to the ca. 780 Ma Laurentia poles \citep{Evans2015a,Fairchild2017a}.  A challenge with this model in terms of regional tectonics, is that it is unclear what conjugate lithosphere rifted from East Greenland leading to the thick sedimentary succession interpreted to have been deposited on a thermally-subsiding passive margin from ca. 850 Ma into the Cryogenian (Fig. \ref{fig:tectonic_history}; \citealp{Maloof2006a}). \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting. The separation of Baltica from Laurentia, once in the configuration shown in Figure \ref{fig:Grenville_reconstructions}, is hypothesized to have occurred ca. 600 Ma associated with the opening of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citealp{Cawood2001a}). The cratons would once again be conjoined ca. 430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.

@@

High-quality paleomagnetic poles constrain the coeval ca. 1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}. These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone. However, with the subsequent rapid motion of Laurentia to low latitudes (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny. This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g. \citealp{Li2008a}). Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift. The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} is that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g. \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia. The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them becoming conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}. The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca. 1090 followed by peak granulite metamorphism ca. 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean. If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca. 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in eastern Laurentia. This late Tonian timing for separation could be  consistent with a position along the southeast Laurentia margin given the ca. 750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Cuyania \citep{Martin2019a}.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the ca. 1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. In terms of departing from this position, one possibility is that its departure is associated with early Neoproterozoic extension in northwest Laurentia.
